Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms

### 2020-03-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Adjust latex style to suit sociological methods and research journal

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},
  pdfauthor={Hanne Oberman},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Missing the Point: Convergence of Multiple Imputation using Chained
Equations Algorithms}
\author{Hanne Oberman}
\date{16-3-2020}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

This note provides the theoretical foundation towards the diagnostic
evaluation of multiple imputation algorithms. For reasons of brevity, we
only focus on the MI algorithm implemented in \texttt{mice} (Van Buuren
and Groothuis-Oudshoorn 2011). The convergence properties of this MI
algorithm are investigated through model-based simulation.\footnote{All
  programming code used in this note is available from
  {[}github.com/gerkovink/shinyMice/simulation{]}\{\url{https://github.com/gerkovink/ShinyMICE/simulation}\}.}
The results of this simulation study are guidelines for assessing
convergence of MI algorithms.

@@

@@

@@

@@

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  Royal, times, sageapa]{sagej}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},
  pdfauthor={Hanne I. Oberman},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Missing the Point: Convergence of Multiple Imputation using Chained
Equations Algorithms}
\author{Hanne I. Oberman}
\date{16-3-2020}

@@

\hypertarget{background}{%
\section{Background}\label{background}}

@@

This note provides the theoretical foundation towards the diagnostic
evaluation of multiple imputation algorithms. For reasons of brevity, we
only focus on the MI algorithm implemented in \texttt{mice} (Van Buuren
and Groothuis-Oudshoorn 2011). The convergence properties of this MI
algorithm are investigated through model-based simulation.\footnote{All
  programming code used in this note is available from
  \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.}
The results of this simulation study are guidelines for assessing
convergence of MI algorithms.

@@

\hypertarget{theoretical-background}{%
\subsubsection{Theoretical Background}\label{theoretical-background}}

There is no scientific consensus on the convergence properties of
multiple imputation algorithms (Takahashi 2017). Some default techniques
in \texttt{mice} might not yield converged states at all (Murray 2018).
Therefore, algorithmic convergence should be monitored carefully. The
diagnostic evaluation of convergence is preferred over the current
practice---visually inspecting imputation chains for signs of
non-convergence---since the latter may be challenging to the untrained
eye (Van Buuren 2018, \textasciitilde6.5.2).

The application of convergence diagnostics to MI methods has not been
systematically studied (Van Buuren 2018). The measures that are
available for diagnosing convergence of iterative algorithmic procedures
(Markov chain Monte Carlo methods; e.g., \texttt{mice} algorithms or
Gibbs samplers) generally evaluate signs of non-convergence (Hoff 2009).
Non-convergence may be present as non-stationarity within chains (i.e.,
trending), or as slow mixing between chains (i.e., no intermingling).

The stationarity component of convergence may be evaluated with
autocorrelation (\(AC\); Schafer 1997; Gelman et al. 2013), numeric
standard error (or `MC error'; Geweke 1992), and Raftery and Lewis's
(1991) procedure to determine the effect of trending within chains.

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
Vehtari et al. (2019), we might also evaluate the stationarity component
of convergence. This would make \(\widehat{R}\) a general convergence
diagnostic. The application of \(\widehat{R}\) to assess stationarity
has not been thoroughly investigated. Therefore, this study employs both
\(\widehat{R}\) and autocorrelation to investigate convergence, as
recommended by (Cowles and Carlin 1996, \textasciitilde898).

The mixing component of convergence may be evaluated with the potential
scale reduction factor, \(\widehat{R}\) (a.k.a. `Gelman-Rubin
statistic'; Gelman and Rubin 1992). The stationarity component may be
assessed through autocorrelations. An adapted version of \(\widehat{R}\)
might also be used to diagnose non-stationarity. Other convergence
diagnostics are outside of the scope of this study.

All of these methods evaluate the convergence of univariate scalar
summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). Van Buuren (2018) proposed
to implement multivariable evaluation through eigenvalue decomposition
(MacKay and Mac Kay 2003). This method is outside of the scope of the
current study (\textbf{not anymore!}).

To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
\textasciitilde5). Let \(M\) be the total number of chains, \(T\) the
number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

\begin{align*}
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
<!-- %\text{ \cite[p.~5]{veht19}}  -->
\end{align*}

We then estimate between-chain variance \(B\) as the variance of the
collection of average \(\theta\) per chain.

\begin{align*}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}. 
<!-- %\text{ \cite[p.~5]{veht19}}  -->
\end{align*}

From the between- and within-chain variances we compute a weighted
average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
total variance of \(\theta\). \(\widehat{R}\) is then obtained as a
ratio between the over-estimated total variance and the within-chain
variance:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{equation*}

We can interpret \(\widehat{R}\) as potential scale reduction factor
since it indicates by how much the variance of \(\theta\) could be
shrunken down if an infinite number of iterations per chain would be run
(Gelman and Rubin 1992). This interpretation assumes that chains are
`over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
Over-dispersion implies that the initial values of the chains are `far
away' from the target distribution and each other. When all chains
sample independent of their initial values, the mixing component of
convergence is satisfied, and \(\widehat{R}\)-values will be close to
one. High \(\widehat{R}\)-values thus indicate non-convergence. The
conventionally acceptable threshold for convergence was
\(\widehat{R} < 1.2\) (Gelman and Rubin 1992). More recently, Vehtari et
al. (2019) proposed a more stringent threshold of
\(\widehat{R} < 1.01\).

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain (Lynch 2007, \textasciitilde147). In this study we only consider
\(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and
\(t+1^{th}\) iteration of the same chain.

\begin{equation*}
AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
\end{equation*}

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Positive \(AC\)-values,
however, indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one-another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
Moreover, the magnitude of \(AC\)-values may be evaluated statistically,
but that is outside of this note's scope.

In short, convergence is reached when there is no dependency between
subsequent iterations of imputation chains (\(AC = 0\)), and chains
intermingle such that the only difference between the chains is caused
by the randomness induced by the algorithm (\(\widehat{R} = 1\)).

\hypertarget{simulation-hypothesis}{%
\subsubsection{Simulation Hypothesis}\label{simulation-hypothesis}}

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods (i.e., average bias, average confidence interval
width, and empirical coverage rate across simulations; Van Buuren 2018,
\textasciitilde2.5.2). That is, there is no baseline measure available
to evaluate performance against.The current practice of visually
inspecting imputation chains for signs of non-mixing or non-stationarity
can only diagnose severely pathological cases of non-convergence. We
could also look at distributional characteristics, and plausibility of
imputed values, see Vink (n.d.). For now, this is outside of the scope
of this study.

Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed (Brooks and Gelman 1998, p\textasciitilde437). In
\texttt{mice}, initial values are chosen randomly from the observed
data. Therefore, we cannot be certain that the initial values are
over-dispersed. We expect this to have little effect on the hypothesized
performance of \(\widehat{R}\). No hypothesis was formulated about the
performance of \(AC\) as convergence diagnostic.

@@

We use model-based simulation in \texttt{R} to perform multiple
imputation under 100 simulation conditions. In each condition, the MI
algorithm comprises of a different imputation chain length
(\(T = 1, 2, \dots, 100\)). The number of simulation runs per condition
is 1000. For each simulation condition (i.e., number of iterations) and
each repetition (i.e., simulation run) we compute convergence
diagnostics (\(\widehat{R}\) and autocorrelation) and simulation
diagnostics (bias, confidence interval width and coverage). These
diagnostics are aggregated across repetitions to obtain their average
value per simulation condition. The simulation set-up is summarized in
the pseudo-code below. The complete \texttt{R} script of the simulation
study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# pseudo-code of simulation }
\NormalTok{simulate data }
\ControlFlowTok{for}\NormalTok{ (number of simulation runs from }\DecValTok{1}\NormalTok{ to }\DecValTok{1000}\NormalTok{)}
  \ControlFlowTok{for}\NormalTok{ (number of iterations from }\DecValTok{1}\NormalTok{ to }\DecValTok{100}\NormalTok{)}
\NormalTok{    create missingness}
\NormalTok{    impute the missingness}
\NormalTok{    compute convergence diagnostics}
\NormalTok{    perform analysis}
\NormalTok{    pool results}
\NormalTok{    compute simulation diagnostics}
\NormalTok{aggregate convergence and simulation diagnostics}
\end{Highlighting}
\end{Shaded}

\hypertarget{data-simulation}{%
\subsubsection{Data Simulation}\label{data-simulation}}

The point of origin for all simulation conditions is a finite population
of \(N=1000\). The data are simulated to solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\):

\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]

The quantity of scientific interest is regression coefficient
\(\beta_1\). The data generating model is a multivariate normal
distribution with the following means structure and variance-covariance
matrix:

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot].
\end{align*}

Data is simulated with the function \texttt{mvtnorm::rmvnorm()}, and
outcome variable \(Y\) is subsequently calculated as
\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\] The complete data is `amputed'
once for each simulation repetition. That is, the function
\texttt{mice::ampute()} is used to impose a missingness mechanism upon
the data. The missingness is univariate, and the probability to be
missing is the same for all four variables, namely 20\%
(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
completely observed. \%The resulting amputed data is equal for all
simulation conditions in the same repetition.

Missing datapoints are imputed with the function \texttt{mice::mice()}.
All MI procedures are performed with Bayesian linear regression
imputation (\texttt{method\ =\ "norm"}), and five imputation chains
(\texttt{m\ =\ 5}). The number of iterations varies between simulation
conditions (\texttt{maxit\ =\ 1,\ 2,\ \textbackslash{}dots,\ 100}).

From the imputation step, we extract chain means and chain variances
across iterations (\(\theta\) at \(t = 1,2,\dots,T\)).\\
We compute convergence diagnostics separately for both \(\theta\)s, and
for each of the four variables in the dataset. We report the maximum
(absolute) value as the `worst linear predictor' of convergence.
\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
recommendations. \(AC\) is computed with function \texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to get variance estimates according
to Rubin's (1987) rules, and subsequently implement finite population
pooling conform Vink and Buuren (2014).

We compute bias as the difference between \(\bar{Q}\) and \(Q\).
Confidence interval width (CIW) is defined as the difference between the
lower and upper bound of the 95\% confidence interval (CI95\%) around
\(\bar{Q}\). We compute the CI95\% bounds as

\[\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},\]

where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. From bias and CIW, we calculate empirical
coverage rates. Coverage rate is the proportion of simulations in which
\(Q\) is between the bounds of the CI95\% around \(\bar{Q}\).

@@

Figures 2 and 3 display results per simulation condition
(\(T = 1,2,\dots,100\)).

\hypertarget{convergence-diagnostics}{%
\subsubsection{Convergence Diagnostics}\label{convergence-diagnostics}}

It is apparent that there is a relation between the number of iterations
per simulation condition (\(T\)) and the convergence diagnostics.
Generally speaking, conditions with longer imputation chains (higher
\(T\)) coincide with less signs of non-convergence
(\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).
We see more or less equivalent trends for chain means versus chain
variances. We, therefore, only discuss the \(\widehat{R}\) and
\(AC\)-values of chain means.

Figure 2A shows that \(\widehat{R}\)-values generally decrease with
increasing imputation chain lengths. The decline stabilizes somewhere
between the simulation conditions \(T=30\) and \(T=50\). The downward
trend is most pronounced for \(T=3\), and between \(T = 5\) and
\(T = 10\). In the intervening conditions (\(3 \leq T \leq 5\)),
however, we observe a steep increase in \(\widehat{R}\)-values. This
increase implies that non-convergence is under-estimated, and
convergence should not be diagnosed at this point. Based on the
conventional threshold \(\widehat{R} < 1.2\), we would falsely diagnose
convergence at \(T=3\). According to the widely used threshold
\(\widehat{R} < 1.1\), convergence would be diagnosed for conditions
where \(T>9\). If we use the recently recommended threshold
\(\widehat{R} < 1.01\), we would conclude that convergence is reached in
none of the simulation conditions.

The \(AC\)-values displayed in Figure 2B are almost uniformly increasing
as a function of \(T\). The only \(AC\)-value that deviates from this
observation is for \(T=2\). The lowest \(AC\)-value is obtained for
\(T=3\). \%At \(T=5\), the \(AC\)-value reaches the level observed at
\(T=2\). The gradual increase plateaus between \(T=10\) and \(T=30\).
\(AC\)-values in conditions where \(T>70\) are indifferentiable from
zero, indicating stationarity. We see an initial decrease between
\(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
\(AC\)-values greater than \(T=2\).

According to this diagnostic, none of the simulation conditions show
signs of non-convergence, since we only observe negative or zero
\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
divergence between subsequent samples in conditions where \(T<4\). It
may not be wise to terminating the algorithm at that point, but after
`recovering' from this dip in \(AC\)-values. This would imply \(T=6\) as
the minimal number of iterations.

Taken together, we see that \(T>3\) is the minimal requirement to
diagnose convergence (\(\widehat{R} < 1.2; AC \leq 0\)). This threshold
is, however, not sufficient, since we overlook the increase in
\(\widehat{R}\)-values up-to conditions where \(T>5\), and the
convergence diagnostics only reach stability at \(T>20\).

\hypertarget{simulation-diagnostics}{%
\subsubsection{Simulation Diagnostics}\label{simulation-diagnostics}}

We use average bias, average confidence interval width, and coverage
rate as performance measures to evaluate \(\widehat{R}\) and \(AC\). We
make the general observation that the simulation diagnostics behave as
theorized for most simulation conditions (bias around zero, stable
confidence interval widths, nominal coverage rate at 95\%).

Figure 3A shows that bias is fairly stable across simulation conditions.
The condition \(T=1\) clearly deviates from this trend with a negative
bias. The bias for \(T=2\) is below average, but within the range of
fluctuations. Conditions where \(T>3\) can be diagnosed as unbiased, but
the average bias across iterations (as shown by a flat Loess line) only
reaches stability at \(T=20\).

CIW is only clearly divergent in the simulation condition \(T=1\), see
Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
is the case for \(T=3\), may yield spurious inferences. Only conditions
where \(T>3\) are therefore considered sufficient. These conditions have
similar CIWs, but across iterations stability is not established (i.e.,
the Loess line is never completely flat within the 100 simulation
conditions). \%After about twenty-five iterations the average CIW across
all simulation conditions and repetitions is reached 0.927.

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% (Neyman 1934), namely 95\%. Similar to
the CIWs, there is some trending across iterations (i.e., the Loess line
is never flat).

From the simulation diagnostics, we observe that unbiased estimates with
nominal coverage rates were obtained in conditions where \(T>3\). This
suggests that as little as four iterations may be sufficient for the MI
algorithm under the current circumstances.

In short, there is a discrepancy between what the convergence
diagnostics and what the performance measures indicate. While \(T=4\)
seems sufficient with respect to simulation quantities, the condition
where \(T=4\) resulted in the `one-but-worst' values of both convergence
diagnostics. Complete algorithmic convergence as indicated by
\(\widehat{R}\) and autocorrelation is not reached in conditions where
\(T<20\).

@@

This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
may diagnose convergence of multiple imputation algorithms, but their
performance differs from conventional applications to iterative
algorithmic procedures.\\
\(\widehat{R}\) and autocorrelation indicate that algorithmic
convergence may only be reached after twenty or even forty iterations,
while unbiased, confidence valid estimates estimates may be obtained
with as little as four iterations. These results are in agreement with
the simulation hypothesis: \(\widehat{R}\) over-estimates the severity
of non-convergence when applied to MI procedures. \%This may be due to
the quantity of scientific interest chosen. More `complicated' \(Q\)s
(e.g., higher order effects or variance components) might show bias,
under- or over-coverage at higher \(T\).

According to this simulation study, the recently proposed threshold of
\(\widehat{R}<1.01\) may be too stringent for MI algorithms. Under the
relatively easy missing data problem of the current study, the threshold
was not reached. The other extreme of the \(\widehat{R}\)-thresholds,
the conventionally acceptable \(\widehat{R} <1.2\), may be too lenient
for MI procedures. Applying this threshold to the current data, lead to
falsely diagnosing convergence at \(T = 3\). It appears that the widely
used threshold of \(\widehat{R} < 1.1\) suits MI algorithms the best. We
might, however, also formulate a new threshold, specifically for the
evaluation of MI algorithms. The current study suggests that
\(\widehat{R} < 1.05\) may be implemented, since that is the level at
which the \(\widehat{R}\) stabilize (around \(T = 20\)).

The negative \(AC\)-values obtained in this study show no threat of
non-stationarity. However, initial dip in \(AC\)-values may have
implications for the default number of iterations in \texttt{mice}
(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
the most appropriate, since this lead to the worst convergence, as
indicated by \(\widehat{R}\) and \(AC\). Under the current
specifications, \(T>20\) would be more appropriate.

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant, since high \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains.
\(AC\) would thus not be informative of the convergence of MI
algorithms.

Since this study only considers only an MCAR missingness mechanism,
results may not be extrapolated to other missing data problems. Proper
performance of the convergence diagnostics under MCAR is necessary but
not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
\(AC\) as convergence diagnostics. This is just a proof of concept.

Further research is needed to investigate their performance under clear
violation of convergence, e.g.~dependency between predictors (predictors
with very high correlations). Until then, we have only shown that the
convergence diagnostics can diagnose non-convergence of MI algorithms
that trend towards a converged state. Also for future research,
implement Wernicke diagnostic, and look at developing a convergence
diagnostic for substantive models, and implement a Wald test for
\(AC = 0\).

@@

\leavevmode\hypertarget{ref-broo98}{}%
Brooks, Stephen P., and Andrew Gelman. 1998. ``General Methods for
Monitoring Convergence of Iterative Simulations.'' \emph{Journal of
Computational and Graphical Statistics} 7 (4): 434--55.
\url{https://doi.org/10.1080/10618600.1998.10474787}.

\leavevmode\hypertarget{ref-cowl96}{}%
Cowles, Mary Kathryn, and Bradley P Carlin. 1996. ``Markov Chain Monte
Carlo Convergence Diagnostics: A Comparative Review.'' \emph{Journal of
the American Statistical Association} 91 (434): 883--904.

\leavevmode\hypertarget{ref-gelm13}{}%
Gelman, Andrew, John B. Carlin, Hal S. Stern, David B. Dunson, Aki
Vehtari, and Donald B. Rubin. 2013. \emph{Bayesian Data Analysis}.
Philadelphia, PA, United States: CRC Press LLC.

\leavevmode\hypertarget{ref-gelm92}{}%
Gelman, Andrew, and Donald B. Rubin. 1992. ``Inference from Iterative
Simulation Using Multiple Sequences.'' \emph{Statistical Science} 7 (4):
457--72. \url{https://doi.org/10.1214/ss/1177011136}.

\leavevmode\hypertarget{ref-gewe92}{}%
Geweke, John. 1992. ``Evaluating the Accuracy of Sampling-Based
Approaches to the Calculations of Posterior Moments.'' \emph{Bayesian
Statistics} 4: 641--49.

\leavevmode\hypertarget{ref-hoff09}{}%
Hoff, Peter D. 2009. \emph{A First Course in Bayesian Statistical
Methods}. Springer Texts in Statistics. New York, NY: Springer New York.
\url{https://doi.org/10.1007/978-0-387-92407-6}.

\leavevmode\hypertarget{ref-lace07}{}%
Lacerda, Miguel, Cally Ardington, and Murray Leibbrandt. 2007.
``Sequential Regression Multiple Imputation for Incomplete Multivariate
Data Using Markov Chain Monte Carlo.'' University of Cape Town, South
Africa.

\leavevmode\hypertarget{ref-lync07}{}%
Lynch, Scott M. 2007. \emph{Introduction to Applied Bayesian Statistics
and Estimation for Social Scientists}. Springer Science \& Business
Media.

\leavevmode\hypertarget{ref-mack03}{}%
MacKay, David JC, and David JC Mac Kay. 2003. \emph{Information Theory,
Inference and Learning Algorithms}. Cambridge university press.

\leavevmode\hypertarget{ref-murr18}{}%
Murray, Jared S. 2018. ``Multiple Imputation: A Review of Practical and
Theoretical Findings.'' \emph{Statistical Science} 33 (2): 142--59.
\url{https://doi.org/10.1214/18-STS644}.
